---
title: "Univariate, Bivariate, and Multivariate Analysis"
author: "Cardiovascular Disease Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 12, fig.height = 6)
```

## Introduction

This notebook contains comprehensive analysis including:
- Univariate Analysis: Analysis of individual variables
- Bivariate Analysis: Analysis of relationships between two variables
- Multivariate Analysis: Analysis of relationships among multiple variables

## Load Libraries and Data

```{r load-libraries}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(dplyr)
library(corrplot)
library(psych)
library(Hmisc)
library(car)
library(gridExtra)
library(factoextra)
library(FactoMineR)

# Set theme for plots
theme_set(theme_minimal() + theme(plot.title = element_text(size = 14, face = "bold")))

# Load the dataset
df <- read.csv("../../data/Cardiovascular_Disease_Dataset.csv", stringsAsFactors = FALSE)

cat("Dataset loaded successfully!\n")
cat("Shape:", nrow(df), "rows,", ncol(df), "columns\n")
head(df)
```

## 1. Univariate Analysis

Univariate analysis involves examining each variable individually to understand its distribution, central tendency, dispersion, and other characteristics.

### 1.1 Univariate Analysis for Numerical Variables

```{r univariate-numerical}
numerical_cols <- c("age", "restingBP", "serumcholestrol", "maxheartrate", "oldpeak")
categorical_cols <- c("gender", "chestpain", "fastingbloodsugar", "restingrelectro", 
                     "exerciseangia", "slope", "noofmajorvessels", "target")

cat(paste(rep("=", 80), collapse = ""), "\n")
cat("UNIVARIATE ANALYSIS - NUMERICAL VARIABLES\n")
cat(paste(rep("=", 80), collapse = ""), "\n")

for (col in numerical_cols) {
  if (col %in% names(df)) {
    cat("\n", toupper(col), ":\n")
    cat("  Count:", sum(!is.na(df[[col]])), "\n")
    cat("  Mean:", mean(df[[col]], na.rm = TRUE), "\n")
    cat("  Median:", median(df[[col]], na.rm = TRUE), "\n")
    cat("  Std Dev:", sd(df[[col]], na.rm = TRUE), "\n")
    cat("  Variance:", var(df[[col]], na.rm = TRUE), "\n")
    cat("  Min:", min(df[[col]], na.rm = TRUE), "\n")
    cat("  Max:", max(df[[col]], na.rm = TRUE), "\n")
    cat("  Range:", max(df[[col]], na.rm = TRUE) - min(df[[col]], na.rm = TRUE), "\n")
    cat("  Q1:", quantile(df[[col]], 0.25, na.rm = TRUE), "\n")
    cat("  Q3:", quantile(df[[col]], 0.75, na.rm = TRUE), "\n")
    cat("  IQR:", IQR(df[[col]], na.rm = TRUE), "\n")
    cat("  Skewness:", psych::skew(df[[col]], na.rm = TRUE), "\n")
    cat("  Kurtosis:", psych::kurtosi(df[[col]], na.rm = TRUE), "\n")
    
    # Outlier detection using IQR method
    Q1 <- quantile(df[[col]], 0.25, na.rm = TRUE)
    Q3 <- quantile(df[[col]], 0.75, na.rm = TRUE)
    IQR_val <- Q3 - Q1
    lower_bound <- Q1 - 1.5 * IQR_val
    upper_bound <- Q3 + 1.5 * IQR_val
    outliers <- df[df[[col]] < lower_bound | df[[col]] > upper_bound, ]
    cat("  Outliers (IQR method):", nrow(outliers), "(", round(nrow(outliers)/nrow(df)*100, 2), "%)\n")
  }
}
```

### 1.2 Visualizations for Univariate Analysis - Numerical Variables

```{r univariate-visualizations}
# Create comprehensive univariate visualizations
plots_list <- list()

for (col in numerical_cols) {
  if (col %in% names(df)) {
    # Histogram
    p1 <- ggplot(df, aes_string(x = col)) +
      geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7, color = "black") +
      geom_vline(aes_string(xintercept = mean(df[[col]], na.rm = TRUE)), 
                 color = "red", linetype = "dashed", size = 1) +
      geom_vline(aes_string(xintercept = median(df[[col]], na.rm = TRUE)), 
                 color = "green", linetype = "dashed", size = 1) +
      labs(title = paste("Histogram:", col),
           x = col,
           y = "Frequency") +
      theme_minimal()
    
    # Box Plot
    p2 <- ggplot(df, aes_string(y = col)) +
      geom_boxplot(fill = "lightblue", alpha = 0.7) +
      labs(title = paste("Box Plot:", col),
           y = col) +
      theme_minimal()
    
    # Q-Q Plot
    p3 <- ggplot(df, aes_string(sample = col)) +
      stat_qq() +
      stat_qq_line() +
      labs(title = paste("Q-Q Plot:", col)) +
      theme_minimal()
    
    plots_list[[paste0(col, "_hist")]] <- p1
    plots_list[[paste0(col, "_box")]] <- p2
    plots_list[[paste0(col, "_qq")]] <- p3
  }
}

# Display first few plots
do.call(grid.arrange, c(plots_list[1:9], ncol = 3))
```

### 1.3 Univariate Analysis for Categorical Variables

```{r univariate-categorical}
cat(paste(rep("=", 80), collapse = ""), "\n")
cat("UNIVARIATE ANALYSIS - CATEGORICAL VARIABLES\n")
cat(paste(rep("=", 80), collapse = ""), "\n")

for (col in categorical_cols) {
  if (col %in% names(df)) {
    cat("\n", toupper(col), ":\n")
    value_counts <- table(df[[col]])
    print(value_counts)
    cat("\n  Percentage distribution:\n")
    for (val in names(value_counts)) {
      cat("    ", val, ":", value_counts[val], "(", round(value_counts[val]/nrow(df)*100, 2), "%)\n")
    }
    cat("  Mode:", names(sort(value_counts, decreasing = TRUE))[1], "\n")
    cat("  Unique values:", length(unique(df[[col]])), "\n")
  }
}
```

### 1.4 Visualizations for Categorical Variables

```{r categorical-visualizations}
categorical_for_vis <- c("gender", "chestpain", "fastingbloodsugar", "exerciseangia", "target")
plots_cat <- list()

for (col in categorical_for_vis) {
  if (col %in% names(df)) {
    value_counts <- table(df[[col]])
    p <- ggplot(df, aes_string(x = paste0("factor(", col, ")"))) +
      geom_bar(fill = "coral", alpha = 0.7, color = "black") +
      labs(title = paste("Bar Chart:", col),
           x = col,
           y = "Count") +
      theme_minimal() +
      geom_text(stat = "count", aes(label = ..count..), vjust = -0.5, size = 3)
    plots_cat[[col]] <- p
  }
}

do.call(grid.arrange, c(plots_cat, ncol = 2))
```

## 2. Bivariate Analysis

Bivariate analysis involves examining the relationship between two variables.

### 2.1 Correlation Analysis - Numerical vs Numerical

```{r correlation-analysis}
cat(paste(rep("=", 80), collapse = ""), "\n")
cat("BIVARIATE ANALYSIS - CORRELATION BETWEEN NUMERICAL VARIABLES\n")
cat(paste(rep("=", 80), collapse = ""), "\n")

correlation_matrix <- cor(df[numerical_cols], use = "complete.obs")
cat("\nCorrelation Matrix:\n")
print(round(correlation_matrix, 3))

# Visualize correlation matrix
corrplot(correlation_matrix, method = "color", type = "upper", 
         order = "hclust", tl.cex = 0.8, tl.col = "black",
         addCoef.col = "black", number.cex = 0.7)
```

### 2.2 Scatter Plots for Numerical Variables vs Target

```{r scatter-plots}
plots_scatter <- list()
for (col in numerical_cols) {
  if (col %in% names(df)) {
    p <- ggplot(df, aes_string(x = col, y = "target", color = "factor(target)")) +
      geom_point(alpha = 0.5) +
      scale_color_manual(values = c("skyblue", "coral"), labels = c("No Disease", "Disease")) +
      labs(title = paste(col, "vs Target"),
           x = col,
           y = "Target",
           color = "Target") +
      theme_minimal()
    plots_scatter[[col]] <- p
  }
}

do.call(grid.arrange, c(plots_scatter, ncol = 2))
```

### 2.3 Numerical vs Categorical (Target) - Group Comparisons

```{r numerical-vs-target}
cat(paste(rep("=", 80), collapse = ""), "\n")
cat("BIVARIATE ANALYSIS - NUMERICAL VARIABLES VS TARGET\n")
cat(paste(rep("=", 80), collapse = ""), "\n")

for (col in numerical_cols) {
  if (col %in% names(df)) {
    cat("\n", toupper(col), "by Target:\n")
    grouped <- df %>%
      group_by(target) %>%
      summarise(
        mean = mean(.data[[col]], na.rm = TRUE),
        median = median(.data[[col]], na.rm = TRUE),
        sd = sd(.data[[col]], na.rm = TRUE),
        min = min(.data[[col]], na.rm = TRUE),
        max = max(.data[[col]], na.rm = TRUE)
      )
    print(grouped)
    
    # Statistical test
    group_0 <- df[df$target == 0, col]
    group_1 <- df[df$target == 1, col]
    test_result <- t.test(group_0, group_1)
    cat("  t-test p-value:", test_result$p.value, 
        ifelse(test_result$p.value < 0.05, "(Significant)", "(Not significant)"), "\n")
  }
}
```

### 2.4 Categorical vs Categorical - Cross-tabulation and Chi-square Tests

```{r categorical-vs-target}
cat(paste(rep("=", 80), collapse = ""), "\n")
cat("BIVARIATE ANALYSIS - CATEGORICAL VARIABLES VS TARGET\n")
cat(paste(rep("=", 80), collapse = ""), "\n")

categorical_vars <- c("gender", "chestpain", "fastingbloodsugar", "restingrelectro", 
                     "exerciseangia", "slope", "noofmajorvessels")

for (var in categorical_vars) {
  if (var %in% names(df)) {
    cat("\n", toupper(var), "vs TARGET:\n")
    crosstab <- table(df[[var]], df$target)
    print(addmargins(crosstab))
    
    # Chi-square test
    contingency_table <- table(df[[var]], df$target)
    test_result <- chisq.test(contingency_table)
    cat("  Chi-square statistic:", test_result$statistic, "\n")
    cat("  p-value:", test_result$p.value, "\n")
    cat("  Significant association:", ifelse(test_result$p.value < 0.05, "Yes", "No"), "(α = 0.05)\n")
    
    # Percentage distribution
    cat("\n  Percentage by Target:\n")
    crosstab_pct <- prop.table(table(df[[var]], df$target), margin = 1) * 100
    print(round(crosstab_pct, 2))
  }
}
```

### 2.5 Violin Plots for Numerical Variables by Target

```{r violin-plots}
plots_violin <- list()
for (col in numerical_cols) {
  if (col %in% names(df)) {
    p <- ggplot(df, aes_string(x = "factor(target)", y = col, fill = "factor(target)")) +
      geom_violin(alpha = 0.7) +
      geom_boxplot(width = 0.1, fill = "white", alpha = 0.5) +
      scale_fill_manual(values = c("skyblue", "coral"), labels = c("No Disease", "Disease")) +
      labs(title = paste(col, "by Target"),
           x = "Target",
           y = col,
           fill = "Target") +
      theme_minimal() +
      theme(legend.position = "none")
    plots_violin[[col]] <- p
  }
}

do.call(grid.arrange, c(plots_violin, ncol = 2))
```

## 3. Multivariate Analysis

Multivariate analysis involves examining relationships among multiple variables simultaneously.

### 3.1 Pairplot for Multivariate Relationships

```{r pairplot}
# Create a pairplot using GGally
library(GGally)

key_vars <- c("age", "restingBP", "maxheartrate", "oldpeak", "target")
sample_df <- df[sample(nrow(df), min(500, nrow(df))), key_vars]
sample_df$target <- factor(sample_df$target)

ggpairs(sample_df, columns = 1:(ncol(sample_df)-1), 
        aes(color = target, alpha = 0.6),
        title = "Pair Plot of Numerical Variables by Target")
```

### 3.2 Principal Component Analysis (PCA)

```{r pca}
cat(paste(rep("=", 80), collapse = ""), "\n")
cat("MULTIVARIATE ANALYSIS - PRINCIPAL COMPONENT ANALYSIS (PCA)\n")
cat(paste(rep("=", 80), collapse = ""), "\n")

# Prepare data for PCA
X <- df[numerical_cols]
X <- X[complete.cases(X), ]
X_scaled <- scale(X)

# Perform PCA
pca_result <- prcomp(X_scaled, center = FALSE, scale. = FALSE)

# Explained variance
explained_variance <- summary(pca_result)$importance[2, ]
cumulative_variance <- summary(pca_result)$importance[3, ]

cat("\nExplained Variance by Component:\n")
for (i in 1:length(explained_variance)) {
  cat("  PC", i, ":", explained_variance[i], "(", explained_variance[i]*100, "%) - Cumulative:", 
      cumulative_variance[i], "(", cumulative_variance[i]*100, "%)\n")
}

# Scree plot
fviz_eig(pca_result, addlabels = TRUE, ylim = c(0, 50))

# Plot first two principal components
X_pca <- as.data.frame(pca_result$x[, 1:2])
X_pca$target <- df[complete.cases(df[numerical_cols]), "target"]

ggplot(X_pca, aes(x = PC1, y = PC2, color = factor(target))) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("skyblue", "coral"), labels = c("No Disease", "Disease")) +
  labs(title = "PCA: First Two Principal Components",
       x = paste0("PC1 (", round(explained_variance[1]*100, 2), "% variance)"),
       y = paste0("PC2 (", round(explained_variance[2]*100, 2), "% variance)"),
       color = "Target") +
  theme_minimal()

cat("\nFirst two principal components explain", cumulative_variance[2]*100, "% of variance\n")
```

### 3.3 Feature Importance Based on Correlation with Target

```{r feature-importance}
cat(paste(rep("=", 80), collapse = ""), "\n")
cat("FEATURE IMPORTANCE - CORRELATION WITH TARGET\n")
cat(paste(rep("=", 80), collapse = ""), "\n")

feature_importance <- data.frame(
  Feature = character(),
  Importance = numeric(),
  stringsAsFactors = FALSE
)

for (col in numerical_cols) {
  if (col %in% names(df)) {
    correlation <- cor(df[[col]], df$target, use = "complete.obs")
    feature_importance <- rbind(feature_importance, 
                               data.frame(Feature = col, Importance = abs(correlation)))
  }
}

# Sort by importance
feature_importance <- feature_importance[order(-feature_importance$Importance), ]

cat("\nFeature Importance (absolute correlation with target):\n")
print(feature_importance)

# Visualize feature importance
ggplot(feature_importance, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  coord_flip() +
  labs(title = "Feature Importance Based on Correlation with Target",
       x = "Feature",
       y = "Absolute Correlation with Target") +
  theme_minimal()
```

### 3.4 Multivariate Correlation Heatmap with Target

```{r multivariate-correlation}
# Multivariate correlation heatmap
numerical_cols_with_target <- c(numerical_cols, "target")
correlation_with_target <- cor(df[numerical_cols_with_target], use = "complete.obs")

corrplot(correlation_with_target, method = "color", type = "upper", 
         order = "hclust", tl.cex = 0.8, tl.col = "black",
         addCoef.col = "black", number.cex = 0.7,
         title = "Multivariate Correlation Matrix (Including Target)")
```

## 4. Key Findings

### Univariate Analysis:
- Distribution characteristics of each variable
- Identification of outliers
- Central tendency and dispersion measures
- Normality assessment

### Bivariate Analysis:
- Correlation between numerical variables
- Relationships between variables and target
- Statistical significance of associations
- Group differences

### Multivariate Analysis:
- Principal component analysis reveals data structure
- Feature importance ranking
- Multivariate relationships and interactions
- Dimensionality reduction insights
